Approximation as Differentiation: Notes

### 2025-05-26 main/higher-order.tex
Tweak proposition 3.3.

$\Fam(\FinVect)$ nor $\Fam(\LatGal)$ provide appropriate notions of composition for automatic differentiation
and Galois slicing respectively, but neither is Cartesian closed. Neither is therefore suitable for
interpreting higher-order programs. However the following is a special case of a construction due to
\citet{nunes2023}:

@@

$\cat{C}$ has all small (set-indexed) products if, for any set $X$ and any $X$-indexed family $A$ of objects
of $\cat{C}$, there exists an object $\prod_{x \in X}A_x$ (also written $\prod_{X} A$) and family of
projection morphisms $\{\eval_x: \prod_{X} A \to A_x\}_{x \in X}$ in $\cat{C}$ such that for any object $B$
and any family of morphisms $\{f_x: B \to A_x\}_{x \in X}$ in $\cat{C}$, there exists a unique product
morphism $\lambda(f)$ making the following diagram commute for every $x \in X$:

@@

If $\cat{C}$ has all small products and also biproducts, then the internal hom $\internalHom{(X, \partial
X)}{(Y, \partial Y)}$ in $\Fam(\cat{C})$ is the object $(Z, \partial Z)$, where:
\begin{itemize}
\item the indexing set $Z$ is the hom-set $\Fam(\cat{C})(X,Y)$, which is indeed a set by
\propref{Grothendieck:fam-inherits};
\item the indexed family $\partial Z$ is the family of objects $\prod_{X}\reindex{\partial Y}{f}$ for every
${(f, \partial f): (X, \partial X) \to (Y, \partial Y)}$.
\end{itemize}

\todo{
\begin{enumerate}
  \item if $\cat{C}$ has biproducts and all small products then $\Fam(\cat{C})$ is Cartesian closed
  \item $\FinVect$ and $\LatGal$ both have biproducts but neither has all small products (\todo{restrict $\LatGal$ to finite lattices?})
  \item $\PSh(\cat{C})$ has all small products (inherited from $\Set$), but doesn't automatically inherit
  biproducts from $\cat{C}$
  \item however, $\PSh_{\CMon}(\cat{C})$ does indeed have biproducts if $\cat{C}$ does
  \item thus $\Fam(\PSh_{\CMon}(\FinVect))$ and $\Fam(\PSh_{\CMon}(\LatGal))$ are both Cartesian closed
  \item also need that $\CMon$-enriched Yoneda embedding preserves biproducts \todo{check: because this is how we will interpret base types? E.g.~$\sem{\mathsf{Bool}} = \Hom{\cat{C}}{-}{2}$ assuming $\cat{C}$ provides a suitable object 2}
  \end{enumerate}
}

@@

\subsubsection{Biproduct laws}
It is also easy to show that in a semi-additive category, as defined in \defref{biproducts:semi-additive}, the
following equations hold:

\vspace{-3mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

\noindent and

@@

Alternatively, we can take these to be the defining equations of a biproduct $(X \biprod Y, \biproj_X,
\biproj_Y, \biinj_X, \biinj_Y)$ in any category $\cat{C}$ known to be $\CMon$-enriched, and therefore
providing addition of morphisms; in that case any products or coproducts in $\cat{C}$ are necessarily
biproducts.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
Suppose $\cat{C}$ is $\CMon$-enriched.
\begin{enumerate}
\item If $(X \times Y, \pi_1, \pi_2)$ is a product in $\cat{C}$ then $(X \times Y, \biinj_X, \biinj_Y, \proj_1, \proj_2)$
is a biproduct, where $\biinj_X = \prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\biinj_Y =
\prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\item If $(X \coprod Y, \inj_1, \inj_2)$ is a coproduct in $\cat{C}$ then $(X \coprod Y, \inj_1, \inj_2,
\biproj_X, \biproj_Y)$ is a biproduct, where $\biproj_X = \coprodM{\id_X}{0_{Y,X}}: X \coprod Y \to X$ and
$\biproj_Y = \coprodM{0_{X,Y}}{\id_Y}: X \coprod Y \to Y$.
\item If $X$ is terminal (or initial) in $\cat{C}$ then it is a zero object.
\end{enumerate}
\end{proposition}

>>>>

$\Fam(\FinVect)$ and $\Fam(\LatGal)$ provide appropriate notions of composition for automatic differentiation
and Galois slicing respectively, but neither is Cartesian closed and thereby a potential candidate for
interpreting higher-order programs. However the following is a special case of a construction due to
\citet{nunes2023}:

@@

$\cat{C}$ has all small (set-indexed) products if, for any set $X$ and any $X$-indexed family $A$ of objects
of $\cat{C}$, there exists an object $\prod_{x \in X}A_x$ (also written $\prod_{X} A$) and family of
projection morphisms $\{\eval_x: \prod_{X} A \to A_x\}_{x \in X}$ in $\cat{C}$ such that for any object $B$
and family of morphisms $\{f_x: B \to A_x\}_{x \in X}$ in $\cat{C}$, there exists a unique product morphism
$\lambda(f)$ making the following diagram commute for every $x \in X$:

@@

If $\cat{C}$ has all small products and also biproducts, then the internal hom $\internalHom{(X, \partial
X)}{(Y, \partial Y)}$ in $\Fam(\cat{C})$ is the object $(Z, \partial Z)$, where:
\begin{itemize}
\item the indexing set $Z$ is the $\Fam(\cat{C})((X, \partial X), (Y, \partial Y))$, which is indeed a set by
\propref{Grothendieck:fam-inherits};
\item the indexed family $\partial Z$ is the family of objects $\prod_{X}\reindex{\partial Y}{f}$ for every
${(f, \partial f): (X, \partial X) \to (Y, \partial Y)}$.
\end{itemize}

\todo{Explain role of biproducts}

\todo{
\begin{enumerate}
  \item $\FinVect$ and $\LatGal$ both have biproducts but neither has all small products (\todo{restrict $\LatGal$ to finite lattices?})
  \item $\PSh(\cat{C})$ has all small products (inherited from $\Set$), but doesn't automatically inherit
  biproducts from $\cat{C}$
  \item however, $\PSh_{\CMon}(\cat{C})$ does indeed have biproducts if $\cat{C}$ does
  \item thus $\Fam(\PSh_{\CMon}(\FinVect))$ and $\Fam(\PSh_{\CMon}(\LatGal))$ are both Cartesian closed
  \item also need that $\CMon$-enriched Yoneda embedding preserves biproducts \todo{check: because this is how we will interpret base types? E.g.~$\sem{\mathsf{Bool}} = \Hom{\cat{C}}{-}{2}$ assuming $\cat{C}$ provides a suitable object 2}
  \end{enumerate}
}

@@

\subsubsection{Biproduct laws}
It is also easy to show that in a semi-additive category the following equations hold:
%\vspace{-4mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

@@

\vspace{3mm}
\noindent It is also possible to use these laws to define biproducts, in the case where $\cat{C}$ is enriched
in $\CMon$ and so equipped with addition of morphisms and zero morphisms. In that case any products or
coproducts in $\cat{C}$ are necessarily biproducts.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
In a category $\cat{C}$ enriched in $\CMon$:
\begin{enumerate}
\item If $(X \times Y, \pi_1, \pi_2)$ is a product then $(X \times Y, \inj_1, \inj_2)$ is a coproduct, where
$\inj_X = \prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\inj_Y = \prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\item If $(X \coprod Y, \inj_1, \inj_2)$ is a coproduct then $(X \coprod Y, \pi_1, \pi_2)$ is a product, where
$\pi_1 = \coprodM{\id_X}{0_{Y,X}}: X \coprod Y \to X$ and $\pi_2 = \coprodM{0_{X,Y}}{\id_Y}: X \coprod Y \to
Y$.
\item If $X$ is terminal (resp.~initial) then $X$ is initial (resp.~terminal).
\end{enumerate}
\end{proposition}
